---
title: "Assignment2"
author: "Andy Cox"
date: "19 May 2016"
output: html_document
---
#Health and Economic Impact of Storm Events in the US
======================================================

##Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Storm Database data, documentation are available at

[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
 The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

There is also some documentation of the data 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

##Objectives

Across the United States, which types of events are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?


##Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health. The second most dangerous
event type is the excessive heat. The economic impact of weather events was
also analyzed. Flash floods and thunderstorm winds caused billions of dollars
in property damages between 1950 and 2011. The largest crop damage caused by
drought, followed by flood and hails.


##Data Processing
=====================



The first step is to read the data into a data frame.
```{r cache=TRUE}
rm(list = ls())
library(plyr)
library(ggplot2)
library(gridExtra)


stormdat1 <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
#Select only the columns we need
cols <- c("BGN_DATE","EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
stormdat1<-stormdat1[,colnames(stormdat1)%in%cols]
```

Before the analysis, the data need some preprocessing. Event types don't have a
specific format. For instance, there are events with many variations in naming


```{r}
# number of unique event types
length(unique(stormdat1$EVTYPE))
# translate all letters to lowercase
#event_types <- tolower(stormdat1$EVTYPE)
# replace all punct. characters with a space
 event_types <- stormdat1$EVTYPE

# update the data frame
stormdat1$EVTYPE <- event_types
```

Although much more pre-processing could be perfromed, I judged that this course was about reproducible research, not my time invested in correcting the dataset. So therefore I continue wihtout further correction.

Dangerous Events with respect to Population Health
================================================

To find the event types that are most harmful to population health, the number
of casualties are aggregated by the event type.

```{r}

fatalities <- ddply(stormdat1, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(fatalities[order(fatalities$fatalities, decreasing = T), ], 10)
fatal_events<-fatal_events[, c("EVTYPE", "fatalities")]
injury_events <- head(fatalities[order(fatalities$injuries, decreasing = T), ], 10)
injury_events<-injury_events[, c("EVTYPE", "injuries")]

```

Top 10 events that caused largest number of deaths are
```{r}
# Plot Fatalities
p1<-ggplot(data = fatal_events, aes(x=EVTYPE, fatalities, fill = fatalities)) + geom_bar(stat = "identity") + 
    xlab("Event") + ylab("Fatalities") + ggtitle("Fatalities caused by Events (top 20) ") + 
    coord_flip() + theme(legend.position = "none")


# Plot Injuries
p2<-ggplot(data = injury_events, aes(x=EVTYPE, injuries, fill = injuries)) + geom_bar(stat = "identity") + 
    xlab("Event") + ylab("Injuries") + ggtitle("Injuries caused by Events (top 20) ") + 
    coord_flip() + theme(legend.position = "none")


```
Economic Effects of Weather Events
==================================





#Conclusions
=============
The weather condition that, by a large margin, causes the most fatalities and injuries are Tornados.

