---
title: "Assignment2"
author: "Andy Cox"
date: "19 May 2016"
output: html_document
---

#Health and Economic Impact of Storm Events in the US

##Synopsis
The analysis on the storm event database revealed that tornadoes are the most
dangerous weather events, casusinf the most numbers of injuries and fatalities.  The economic impact of weather events was also analyzed. The largest number of property damages are casued by tornadoes, whilst the largest crop damage is casued by hail.

##Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Storm Database data, documentation are available at

[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
 The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

There is also some documentation of the data 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

##Objectives
Across the United States, which types of events are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

##Data Processing
The working environment is cleared and required packages are loaded. 
```{r cache=TRUE}
rm(list = ls())
library(ggplot2)
library(gridExtra)
library(plyr)
library(stringr)
```
Then data is laoded directly from the zip file. To save space in memory only the variables needed for the analysis are loaded.

```{r}
stormdat1 <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
#Select only the columns we need
cols <- c("BGN_DATE","EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
stormdat1<-stormdat1[,colnames(stormdat1)%in%cols]
```
**Justification for Data Transformations**

Before the analysis, the data need some preprocessing. Event types have over 900 levels and those levels don't have a specific and clear format. For instance, there are events with many variations in naming conventions. Intial Inspection shows alot of work is needed to correct this. However, as this assigment is about reproducible research and not data analysis, only a minimum will be done to transfrom the data

```{r}
# number of unique event types
length(unique(stormdat1$EVTYPE))

 event_types <- stormdat1$EVTYPE
# translate all letters to uppercase
event_types <- toupper(str_trim(event_types))
event_types <- gsub("TSTM WIND", "MARINE THUNDERSTORM WIND" , event_types)
event_types <- gsub("URBAN/SML STREAM FLD", "HEAVY RAIN", event_types)
event_types <- gsub("MARINE TSTM WIND","MARINE THUNDERSTORM WIND", event_types)
event_types <- gsub("WILD/FOREST FIRE", "WILDFIRE", event_types)
event_types <- gsub("marinethunderstormwind/hail", "marinethunderstormwind", event_types)
event_types <- gsub("TSTM WIND/HAIL","MARINE THUNDERSTORM WIND", event_types)
event_types <- gsub("flashflooding", "flashflood", event_types)
event_types <- gsub("FLOOD/FLASH FLOOD", "FLASH FLOOD", event_types)
event_types <- gsub("WINTER data/MIX", "WINTER data", event_types)
event_types <- gsub("RIP CURRENTS", "RIP CURRENT", event_types)
event_types <- gsub("DENSEDENSEFOG", "DENSE FOG", event_types)
event_types <- gsub("STRONG WINDS","ASTROMICAL LOW TIDE", event_types)
event_types <- gsub("COASTAL FLOODING", "COASTAL FLOOD ", event_types)
event_types <- gsub("RIVER FLOOD", "FLOOD", event_types)
event_types <- gsub("RECORD WARMTH", "HEAT", event_types)
event_types <- gsub("RECORD HEAT", "HEAT", event_types)
event_types <- gsub("FREEZE", "FROST/FREEZE", event_types)
event_types <- gsub("HEATWAVE", "EXCESSIVE HEAT", event_types)
event_types <- gsub("HURRICANE/TYPHOON", "HURRICANE", event_types)

# replace all punct. characters with a space
# update the data frame
stormdat1$EVTYPE <- event_types
```

#Results
##Dangerous Events with respect to Population Health

To find the event types that are most harmful to population health, the number
of casualties and injuries are aggregated by the event type.

```{r}
fatalities <- ddply(stormdat1, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(fatalities[order(fatalities$fatalities, decreasing = T), ], 10)
fatal_events<-fatal_events[, c("EVTYPE", "fatalities")]
fatal_events
injury_events <- head(fatalities[order(fatalities$injuries, decreasing = T), ], 10)
injury_events<-injury_events[, c("EVTYPE", "injuries")]
injury_events
```

Top 10 events that caused largest number of deaths are

**Figure 1.** Barplot showing the frequencies of fatalities due to specific weather events.
```{r echo=FALSE}
# Plot Fatalities
ggplot(data = fatal_events, aes(x=EVTYPE, fatalities, fill = fatalities)) + geom_bar(stat = "identity") + 
    xlab("Event") + ylab("Fatalities") + ggtitle("Fatalities caused by Events (top 10) ") + 
    coord_flip() + theme(legend.position = "none")
```

**Figure 2.** Barplot showing the frequencies of injuries due to specific weather events.
```{r echo=FALSE}
# Plot Injuries
ggplot(data = injury_events, aes(x=EVTYPE, injuries, fill = injuries)) + geom_bar(stat = "identity") + 
    xlab("Event") + ylab("Injuries") + ggtitle("Injuries caused by Events (top 10) ") + 
    coord_flip() + theme(legend.position = "none")
```
#Economic Effects of Weather Events
To find the event types that are associated wiht the greatest economic loss, the  losses are aggregated by the event type.

```{r}
econ_loss <- ddply(stormdat1, .(EVTYPE), summarize,
                    property = sum(PROPDMG),
                    crop = sum(CROPDMG))

# Find events that caused most death and injury
prop_events <- head(econ_loss[order(econ_loss$property, decreasing = T), ], 10)
prop_events<-prop_events[, c("EVTYPE", "property")]
prop_events
crop_events <- head(econ_loss[order(econ_loss$crop, decreasing = T), ], 10)
crop_events<-crop_events[, c("EVTYPE", "crop")]
crop_events
```

**Figure 3.** Barplot showing the economic losses due to property related events
```{r echo=FALSE}
# Plot Economic Losses due to property damages
ggplot(data = prop_events, aes(x=EVTYPE, property, fill = property)) + geom_bar(stat = "identity") + 
    xlab("Event") + ylab("Property Damage") + ggtitle("Property Damage caused by Events (top 10) ") + 
    coord_flip() + theme(legend.position = "none")
```

#Conclusions
The weather condition that, by a large margin, causes the most fatalities and injuries are Tornados. Tornado's also cause the most proporty damages, whilst most crop damages are casued by hail.

